Perturbing SU($n$)-symmetric interactions

### 2020-01-13 collective_sun/sun_perturbation.tex
discuss the point of partitions in Appendix D, improve a bit of notation around e.g. v vs v_\Delta

We wish to determine the effective dynamics induced on the
ground-state manifold $\M_0$ of $H_0$ by weak single- and two-body
perturbations of the form
\begin{align}
  \V_1 \equiv \sum_p v_p X_p = \v v\c\v X,
  &&
  \V_2 \equiv \sum_{p<q} w_{pq} \times \f12 \p{X_p Y_q + X_q Y_p}
  = \f12 \v X \c\m w \c\v Y,
  \label{eq:perturbations}
\end{align}
where $X,Y$ are self-adjoint\footnote{In future versions of these
  notes, we should relax the restriction that $X,Y$ are self-adjoint.}
trace-zero operators on an $n$-dimensional Hilbert space; $X_p,Y_p$
denote the action of $X,Y$ on spin $p$; $\v X,$ $\v Y$, $\v v$, and
$\m w$ are respectively vectors and a matrix of all
$X_p,Y_p,v_p,w_{pq}$; we use a dot ($\c$) to denote an inner product
between vectors and matrices as appropriate; and we enforce
$w_{pq}=w_{qp}$ and $w_{pp}=0$.

If the coefficients $h_{pq}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of the interaction
Hamiltonian $H_0$ consists of fully symmetric states that are
simultaneous $+1$ eigenstates of all permutation operators $\Pi_{pq}$.
We will adopt the restriction that all $h_{pq}<0$ throughout these
notes, but note that this restriction can be relaxed to the assumption
that the initial state of the spins is fully symmetric (e.g.~a
spin-polarized state).  We will also assume throughout these notes
that the fully symmetric manifold $\M_0$ is gapped away from all other
states by an interaction energy difference that is large compared to
any coupling between $\M_0$ and its orthogonal complement.

The effective Hamiltonian $H_M$ induced on the ground-state manifold
$\M_0$ by an $M$-body perturbation $\V_M$ through second order in
perturbation theory is given by\cite{bravyi2011schrieffer,
  perlin2019effective}
\begin{align}
  H_M = H_M^{(1)} + H_M^{(2)},
  &&
  H_M^{(1)} = \P_0 \V_M \P_0,
  &&
  H_M^{(2)} = - \P_0 \V_M \E \V_M \P_0,
  &&
  \E \equiv \sum_{\Delta>0} \f{\P_\Delta}{\Delta},
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace of the
interaction Hamiltonian $H_0$, with interaction energy $\Delta$ above
that of fully symmetric manifold $\M_0$; that is, the interaction
energy of states in the image of $\P_\Delta$ is $E_0+\Delta$, where
\begin{align}
  E_0 = \sum_{p<q} h_{pq}
\end{align}
is the interaction energy of fully symmetric states $\ket\psi\in\M_0$.

@@

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_p v_p \P_0 X_p \P_0
  = \sum_p v_p \P_0 X_0 \P_0 \EQFS \bar v \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v \equiv \f1N \sum_p v_p,
  &&
  \col{X} \equiv \sum_p X_p.
\end{align}
In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\sum_{p,q} h_{pq} v_p X_q - \sum_q h_q v_q X_q} \ket\psi,
  \label{eq:H_V1_psi}
\end{align}
where $h_q\equiv\sum_p h_{pq}$ is the net interaction energy
associated with spin $p$.  We thus find that the vector $\V_1\ket\psi$
is an eigenvector of the interaction Hamiltonian $H_0$ if the
coefficients $\v v$ satisfy
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta\p{\v v} \v v,
  \label{eq:cond_1}
\end{align}
where $\v h,\m h$ are respectively a vector and matrix of all
$h_p,h_{pq}$; $\diag\v h$ is a matrix with $\v h$ on the diagonal and
zeros everywhere else; and $\Delta\p{\v v}$ is some constant that may
depend on $\v v$.  Solving the general eigenvalue problem in
\eqref{eq:cond_1} requires diagonalizing an $N\times N$ matrix.  When
the interaction Hamiltonian $H_0$ is translationally invariant,
however, this eigenvalue problem can be solved analytically; we
provide the corresponding solution in Appendix \ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_1\p{\v v_\Delta}$ with
coefficients $\v v_\Delta$ that satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_1\p{\v v_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_1\p{\v v_\Delta} \ket\psi.
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1\p{\v v_\Delta}\ket\psi$ depends only on the coefficients
$\v v_\Delta$, and is entirely independent of the state $\ket\psi$ or
choice of single-spin trace-zero operator $X$ used to build $\V_1$ in
\eqref{eq:perturbations}.  Finding operators $\V_1\p{\v v_\Delta}$
that generate eigenvectors of the interaction Hamiltonian $H_0$ when
they act on fully symmetric states $\ket\psi$ thus reduces to finding
eigenvectors $\v v_\Delta$ of the interaction matrix $\m h-\diag\v h$.

@@

We now return to the task of computing the second-order effective
Hamiltonian $H_1^{(2)}$.  Any coefficient vector $\v v$ can be
expanded into its projections $\v v_\Delta$ onto the eigenspace of
vectors satisfying \eqref{eq:cond_1} with eigenvalue $\Delta$,
i.e.~$\v v = \sum_\Delta \v v_\Delta$, which also allows us to expand
\begin{align}
  \V_1\p{\v v} = \sum_\Delta \V_1\p{\v v_\Delta},
\end{align}
where each operator $\V_1\p{\v v_\Delta}$ generates a state with
definite interaction energy $\Delta$ above that of the fully symmetric
manifold $\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1\p{\v v} \E \V_1\p{\v v} \P_0
  = - \sum_{\Delta>0} \f1{\Delta}
  \P_0 \V_1\p{\v v} \P_\Delta \V_1\p{\v v} \P_0
  = - \sum_{\Delta>0} \f1{\Delta} \P_0 \V_1\p{\v v_\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \V_1\p{\v v_\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PXXP}.  In total, we find
\begin{align}
  H_1^{(2)}
  \EQFS \f1{N\p{N-1}}
  \sum_{\Delta>0} \f{\v v_\Delta\c\v v_\Delta}{\Delta}
  \sp{\p{\col{X}}^2 - N \p{\col{X^2}}},
  \label{eq:H_1_2}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, $\v v_\Delta$ is the projection of $\v v$
onto the $\Delta$-eigenspace of the matrix $\m h-\diag\v h$, and
\begin{align}
  \col{X} \equiv \sum_p X_p,
  &&
  \col{X^2} \equiv \sum_p X_p^2.
\end{align}

@@

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$ (see Appendix \ref{sec:PXYP}):
\begin{align}
  H_2^{(1)}
  = \P_0 \V_2 \P_0
  \EQFS \f12 \bar w \p{\col{X}\odot\col{Y} - \col{X\odot Y}},
\end{align}
where $A\odot B\equiv\p{AB+BA}/2$ is a symmetrized product of $A$ and
$B$; and
\begin{align}
  \bar w \equiv {N\choose 2}^{-1} \sum_{p<q} w_{pq},
  &&
  \col{X} \equiv \sum_p X_p,
  &&
  \col{Y} \equiv \sum_p Y_p,
  &&
  \col{X\odot Y} \equiv \sum_p X_p\odot Y_p.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \f12 \sum_{p\ne q} \sp{2\bar h_{pq} w_{pq}
    + \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}}
  X_p Y_q \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq}.
\end{align}
We thus find that the vector $\V_2\ket\psi$ is an eigenvector of the
interaction Hamiltonian $H_0$ if the coefficients $\m w$ satisfy
\begin{align}
  2 \bar h_{pq} w_{pq}
  + \sum_{k\notin\set{p,q}} \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  = \Delta\p{\m w} w_{pq},
  \label{eq:cond_2}
\end{align}
where $\Delta\p{\m w}$ is some constant that may depend on $\m w$, and
for clarity we explicitly exclude terms with $k\in\set{p,q}$ that
vanish due to the fact that $\m h$ and $\m w$ are zero on the
diagonal.  We can recast the condition in \eqref{eq:cond_2} as an
eigenvalue problem by writing the coefficient matrix $\m w$ as a
vector $\v{\m w}\in\RR^{N\choose 2}$,
\begin{align}
  \m w = \sum_{p,q} w_{pq} \op{p}{q}
  \to \v{\m w} \equiv \flat\m w
  \equiv \sum_{p<q} w_{pq} \ket{\set{p,q}},
\end{align}
where we label basis of vectors $\ket{\set{p,q}}\in\RR^{N\choose 2}$
by a choice of two distinct integers $p,q\in\ZZ_N$.  We similarly
define the ${N \choose 2}\times{N \choose 2}$ matrix
\begin{align}
  \check{\m h} = 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
The condition in \eqref{eq:cond_2} is then equivalently the eigenvalue
equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta\p{\m w} \v{\m w}.
  \label{eq:cond_2_eig}
\end{align}
When the interaction Hamiltonian $H_0$ and perturbation $\V_2\p{\m w}$
are translationally invariant, this eigenvalue problem can be reduced
to that of diagonalizing a matrix of dimensions
$\floor{N/2}\times\floor{N/2}$, as we show in Appendix
\ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_2\p{\m w_\Delta}$ with
coefficients $\m w_\Delta$ that satisfy \eqref{eq:cond_2_eig} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_2\p{\m w_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_2\p{\m w_\Delta} \ket\psi,
\end{align}
which implies that $\V_2\p{\m w_\Delta}\ket\psi$ is a state of
definite energy that depends only on the coefficients $\m w_\Delta$.
As with the case of single-body perturbations, we can decompose any
matrix $\m w$ into its projections $\m w_\Delta$ onto the eigenspace
of matrices satisfying \eqref{eq:cond_2_eig} with eigenvalue $\Delta$,
i.e.~$\m w=\sum_\Delta\m w_\Delta$, which allows us to expand
\begin{align}
  \V_2\p{\m w} = \sum_\Delta \V_2\p{\m w_\Delta},
\end{align}
and in turn write
\begin{align}
  H_2^{(2)} = - \P_0 \V_2\p{\m w} \E \V_2\p{\m w} \P_0
  = -\sum_{\Delta>0} \f1\Delta \P_0 \V_2\p{\m w_\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \V_2\p{\m w_\Delta}^2 \P_0$ is simplified in
Appendices \ref{sec:PXYXYP} and \ref{sec:multi_to_collective}.  If
$X=Y=s_\z$, where $s_\z\equiv\p{\op\up-\op\dn}/2$ is an SU(2) spin
operator, then we can expand
\begin{align}
  H_2^{(2)} \EQFS W_4\p{\m h,\m w} S_\z^4 - W_2\p{\m h,\m w} S_\z^2,
\end{align}
where $S_\z\equiv\sum_p s_\z^{(p)}$ is a collective SU(2) spin
operator, $W_k\p{\m h,\m w}$ are scalar coefficients, and we have
neglected scalar terms in $H_2^{(2)}$ without physical consequence.
In order to expand the coefficients $W_k\p{\m h,\m w}$, for any
symmetric matrix $\m m$ with zeroes on the diagonal we define the
vectors
\begin{align}
  \v{\m m} \equiv \flat\m m \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  &&
  \v m \equiv \sum_{p,q} m_{pq} \ket{p},
\end{align}
in terms of which
\begin{align}
  W_4\p{\m h,\m w}
  \equiv \sum_\Delta \f1\Delta \f{\v w_\Delta\c\v w_\Delta
    - \v{\m w}_\Delta\c\v{\m w}_\Delta}{N!_4},
\end{align}
\begin{align}
  W_2\p{\m h,\m w}
  \equiv \sum_\Delta \f1{\Delta} \sp{\f{\v w_\Delta\c\v w_\Delta
      - 2\v{\m w}_\Delta\c\v{\m w}_\Delta}{4N!_2}
    + \f{\p{3N-4}\p{\v w_\Delta\c\v w_\Delta
        - \v{\m w}_\Delta\c\v{\m w}_\Delta}}{2N!_4}},
\end{align}
where
\begin{align}
  N!_k \equiv \prod_{j=0}^{k-1} \p{N-j} = \f{N!}{\p{N-k}!}.
\end{align}
Note that the dependence of the coefficients of $W_k\p{\m h,\m w}$ on
the interaction matrix $\m h$ is implicit in the decomposition of the
coefficient matrix $\m w$ into components $\m w_\Delta$ as
$\m w=\sum_\Delta\m w_\Delta$, where $\m w_\Delta$ is the projection
of $\m w$ onto the $\Delta$-eigenspace of the operator $\check{\m h}$
defined in \eqref{eq:h_super_mat}.

@@

We wish to simplify
\begin{align}
  H_0 \V_2 \ket\psi
  = \f12 \sum_{\substack{k,\ell\\p<q}} w_{k\ell} h_{pq}
  \Pi_{pq} X_k Y_\ell \ket\psi,
  \label{eq:H_V2_psi_start}
\end{align}
which has one term with $\set{p,q}=\set{k,\ell}$, terms with
$p,q\notin\set{k,\ell}$, and terms with only one of
$p,q\in\set{k,\ell}$.  The permutation operator $\Pi_{pq}$ acts
trivially on $X_k Y_\ell \ket\psi$ when $\set{p,q}=\set{k,\ell}$ or
$p,q\notin\set{k,\ell}$, leaving a sum over $p,q$ of the form
\begin{align}
  h_{k\ell} + \sum_{\substack{p<q\\p,q\notin\set{k,\ell}}} h_{pq}
  = \sum_{p<q} h_{pq}
  - \sum_{\substack{p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  = E_0 + 2 h_{k\ell} - h_k - h_\ell
  = E_0 + 2 \bar h_{k\ell},
\end{align}
where
\begin{align}
  \bar h_{k\ell} \equiv h_{k\ell} - \f12\p{h_k + h_\ell}.
\end{align}
The corresponding terms in \eqref{eq:H_V2_psi_start} with
$p,q\notin\set{k,\ell}$ and $\set{p,q}=\set{k,\ell}$ are then
\begin{align}
  E_0 \V_2 \ket\psi
  + \sum_{k,\ell} \bar h_{k\ell} w_{k\ell} X_k Y_\ell \ket\psi.
\end{align}
Using similar techniques, we can simplify the terms in
\eqref{eq:H_V2_psi_start} with only one of $p,q\in\set{k,\ell}$ to
\begin{align}
  \f12 \sum_{k\ne\ell} \sum_q \p{h_{kq} w_{q\ell} + w_{kq} h_{q\ell}}
  X_k Y_\ell \ket\psi.
\end{align}
In total,
\begin{align}
  H_0 \V_2 \ket\psi
  &= E_0 \V_2 \ket\psi
  + \sum_{k,\ell} \bar h_{k\ell} w_{k\ell} X_k Y_\ell \ket\psi
  + \f12 \sum_{k\ne\ell} \sum_q \p{h_{kq} w_{q\ell} + w_{kq} h_{q\ell}}
  X_k Y_\ell \ket\psi \\
  &= E_0 \V_2 \ket\psi
  + \f12 \sum_{k\ne\ell} \sp{2\bar h_{k\ell} w_{k\ell}
    + \sum_q \p{h_{kq} w_{q\ell} + w_{kq} h_{q\ell}}}
  X_k Y_\ell \ket\psi
\end{align}

>>>>

We wish to determine the effective dynamics induced on the
ground-state manifold $\M_0$ of $H_0$ by weak single- and two-body
perturbations of the form
\begin{align}
  \V_1 \equiv \sum_p v_p X_p = \v v\c\v X,
  &&
  \V_2 \equiv \sum_{p<q} w_{pq} \times \f12 \p{X_p Y_q + X_q Y_p}
  = \f12 \v X \c\m w \c\v Y,
  \label{eq:perturbations}
\end{align}
where $X,Y$ are self-adjoint\footnote{In future versions of these
  notes, we should relax the restriction that $X,Y$ are self-adjoint.}
trace-zero operators on an $n$-dimensional Hilbert space; $X_p,Y_p$
denote the action of $X,Y$ on spin $p$; $\v X,$ $\v Y$, $\v v$, and
$\m w$ are respectively vectors and a matrix of all
$X_p,Y_p,v_p,w_{pq}$; we use a dot ($\c$) to denote an inner product
between vectors and matrices as appropriate; and we enforce
$w_{pq}=w_{qp}$ and $w_{pp}=0$.
% TODO: consider more general perturbations, namely:
% (1) a sum over X in V_1,
% (2) more general two-body perturbations (O_pq) in V_2, and
% (3) a sum over different O_pq in V_2

If the coefficients $h_{pq}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of the interaction
Hamiltonian $H_0$ consists of fully symmetric states that are
simultaneous $+1$ eigenstates of all permutation operators $\Pi_{pq}$.
We will adopt the restriction that all $h_{pq}<0$ throughout these
notes, but note that this restriction can be relaxed to the assumption
that the initial state of the spins is fully symmetric (e.g.~a
spin-polarized state).  We will also assume throughout these notes
that the fully symmetric manifold $\M_0$ is gapped away from all other
states by an interaction energy difference that is large compared to
any coupling between $\M_0$ and its orthogonal complement.  The
precise conditions for such a non-vanishing many-body interaction
energy gap is an important question that deserves further study in
future work.

The effective Hamiltonian $H_M$ induced on the ground-state manifold
$\M_0$ by an $M$-body perturbation $\V_M$ through second order in
perturbation theory is given by\cite{bravyi2011schrieffer,
  perlin2019effective}
\begin{align}
  H_M = H_M^{(1)} + H_M^{(2)},
  &&
  H_M^{(1)} = \P_0 \V_M \P_0,
  &&
  H_M^{(2)} = - \P_0 \V_M \E \V_M \P_0,
  &&
  \E \equiv \sum_{\Delta\ne0} \f{\P_\Delta}{\Delta},
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace of the
interaction Hamiltonian $H_0$, with interaction energy $\Delta$ above
that of fully symmetric manifold $\M_0$; that is, the interaction
energy of states in the image of $\P_\Delta$ is $E_0+\Delta$, where
\begin{align}
  E_0 = \sum_{p<q} h_{pq}
\end{align}
is the interaction energy of fully symmetric states $\ket\psi\in\M_0$.

@@

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_p v_p \P_0 X_p \P_0
  = \sum_p v_p \P_0 X_0 \P_0 \EQFS \bar v \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v \equiv \f1N \sum_p v_p,
  &&
  \col{X} \equiv \sum_p X_p.
\end{align}
In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1\p{\v v} \ket\psi
  = E_0 \V_1\p{\v v} \ket\psi
  + \p{\sum_{p,q} h_{pq} v_p X_q - \sum_q h_q v_q X_q} \ket\psi,
  &&
  h_q \equiv \sum_p h_{pq}.
  \label{eq:H_V1_psi}
\end{align}
We thus find that $\V_1\p{\v v_\Delta}\ket\psi$ is an eigenvector of
the interaction Hamiltonian $H_0$ if the coefficients $\v v_\Delta$
satisfy
\begin{align}
  \p{\m h - \diag\v h}\c\v v_\Delta = \Delta \v v_\Delta,
  \label{eq:cond_1}
\end{align}
where $\v h,\m h$ are respectively a vector and matrix of all
$h_p,h_{pq}$; $\diag\v h$ is a matrix with $\v h$ on the diagonal and
zeros everywhere else; and $\Delta$ is a scalar.  Solving the general
eigenvalue problem in \eqref{eq:cond_1} requires diagonalizing an
$N\times N$ matrix.  When the interaction Hamiltonian $H_0$ is
translationally invariant, however, this eigenvalue problem can be
solved analytically; we provide the corresponding solution in Appendix
\ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_1\p{\v v_\Delta}$ with
coefficients $\v v_\Delta$ that satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_1\p{\v v_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_1\p{\v v_\Delta} \ket\psi.
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1\p{\v v_\Delta}\ket\psi$ depends only on the coefficients
$\v v_\Delta$, and is entirely independent of the state $\ket\psi$ or
the choice of a single-spin trace-zero operator $X$ used to build
$\V_1$ in \eqref{eq:perturbations}.
% TODO: consider more general single-body perturbations V_1, which sum
% over different operators X
Finding operators $\V_1\p{\v v_\Delta}$ that generate eigenvectors of
the interaction Hamiltonian $H_0$ when they act on fully symmetric
states $\ket\psi$ thus reduces to finding eigenvectors $\v v_\Delta$
of the interaction matrix $\m h-\diag\v h$.

@@

We now return to the task of computing the second-order effective
Hamiltonian $H_1^{(2)}$.  Any coefficient vector $\v v$ can be
expanded into its projections $\v v_\Delta$ onto the eigenspace of
vectors satisfying \eqref{eq:cond_1} with eigenvalue $\Delta$,
i.e.~$\v v = \sum_\Delta \v v_\Delta$, which also allows us to expand
\begin{align}
  \V_1\p{\v v} = \sum_\Delta \V_1\p{\v v_\Delta},
\end{align}
where each operator $\V_1\p{\v v_\Delta}$ generates a state with
definite interaction energy $\Delta$ above that of the fully symmetric
manifold $\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1\p{\v v} \E \V_1\p{\v v} \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta}
  \P_0 \V_1\p{\v v} \P_\Delta \V_1\p{\v v} \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta} \P_0 \V_1\p{\v v_\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \V_1\p{\v v_\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PXXP}.  In total, we find
\begin{align}
  H_1^{(2)}
  \EQFS \f1{N\p{N-1}}
  \sum_{\Delta\ne0} \f{\v v_\Delta\c\v v_\Delta}{\Delta}
  \sp{{\col{X}\,}^2 - N \col{X^2}},
  \label{eq:H_1_2}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, $\v v_\Delta$ is the projection of $\v v$
onto the $\Delta$-eigenspace of the matrix $\m h-\diag\v h$, and
\begin{align}
  \col{X} \equiv \sum_p X_p,
  &&
  \col{X^2} \equiv \sum_p X_p^2.
\end{align}

@@

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$ (see Appendix \ref{sec:PXYP}):
\begin{align}
  H_2^{(1)}
  = \P_0 \V_2 \P_0
  \EQFS \f12 \bar w \p{\col{X}\odot\col{Y} - \col{X\odot Y}},
\end{align}
where $A\odot B\equiv\p{AB+BA}/2$ is a symmetrized product of $A$ and
$B$; and
\begin{align}
  \bar w \equiv {N\choose 2}^{-1} \sum_{p<q} w_{pq},
  &&
  \col{X} \equiv \sum_p X_p,
  &&
  \col{Y} \equiv \sum_p Y_p,
  &&
  \col{X\odot Y} \equiv \sum_p X_p\odot Y_p.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2\p{\m w} \ket\psi
  = E_0 \V_2\p{\m w} \ket\psi
  + \f12 \sum_{p\ne q} \sp{2\bar h_{pq} w_{pq}
    + \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}}
  X_p Y_q \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq}.
\end{align}
We thus find that the vector $\V_2\p{\m w_\Delta}\ket\psi$ is an
eigenvector of the interaction Hamiltonian $H_0$ if the coefficients
$\m w_\Delta$ satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w}_\Delta = \Delta \v{\m w}_\Delta,
  \label{eq:cond_2}
\end{align}
where for any symmetric $N\times N$ matrix $\m m$ with zeroes on the
diagonal (as is the case with the coefficient matrix $\m w_\Delta$),
we define the ${N \choose 2}$-component vector
\begin{align}
  \v{\m m} \equiv \sum_{p<q} m_{pq} \ket{\set{p,q}},
  \label{eq:flat_vec}
\end{align}
and we define the ${N \choose 2} \times {N \choose 2}$ matrix
\begin{align}
  \check{\m h} = 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
The basis vectors $\ket{\set{p,q}}$ are labeled by a choice
$\set{p,q}$ of two distinct spins $p,q$.  The eigenvalue problem in
\eqref{eq:cond_2} can be simplified when the interaction Hamiltonian
$H_0$ and the perturbation $\V_2\p{\m w}$ are translationally
invariant; we discuss this simplification in Appendix
\ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_2\p{\m w_\Delta}$ with
coefficients $\m w_\Delta$ that satisfy \eqref{eq:cond_2} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_2\p{\m w_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_2\p{\m w_\Delta} \ket\psi,
\end{align}
which implies that $\V_2\p{\m w_\Delta}\ket\psi$ is a state of
definite energy that depends only on the coefficients $\m w_\Delta$.
As with the case of single-body perturbations, we can decompose any
matrix $\m w$ into its projections $\m w_\Delta$ onto the eigenspace
of matrices satisfying \eqref{eq:cond_2} with eigenvalue $\Delta$,
i.e.~$\m w=\sum_\Delta\m w_\Delta$, which allows us to expand
\begin{align}
  \V_2\p{\m w} = \sum_\Delta \V_2\p{\m w_\Delta},
\end{align}
and in turn write
\begin{align}
  H_2^{(2)} = - \P_0 \V_2\p{\m w} \E \V_2\p{\m w} \P_0
  = -\sum_{\Delta\ne0} \f1\Delta \P_0 \V_2\p{\m w_\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \V_2\p{\m w_\Delta}^2 \P_0$ is simplified in
Appendices \ref{sec:PXYXYP} and \ref{sec:multi_to_collective}.  If
$X=Y=s_\z$, where $s_\z\equiv\p{\op\up-\op\dn}/2$ is an SU(2) spin
operator, then
\begin{align}
  H_2^{(2)} \EQFS W_4\p{\m h,\m w} S_\z^4 - W_2\p{\m h,\m w} S_\z^2,
\end{align}
where $S_\z\equiv\sum_p s_\z^{(p)}$ is a collective SU(2) spin
operator, $W_k\p{\m h,\m w}$ are scalar coefficients, and we have
neglected scalar terms in $H_2^{(2)}$ without physical consequence.
In order to expand the coefficients $W_k\p{\m h,\m w}$, for any
symmetric matrix $\m m$ with zeros on the diagonal, similarly to the
special definition $\v{\m m}$ in \eqref{eq:flat_vec} we define
\begin{align}
  \v m \equiv \sum_{p,q} m_{pq} \ket{p},
\end{align}
which is essentially the sum over all columns (or rows) of $\m m$, in
terms of which
\begin{align}
  W_4\p{\m h,\m w}
  \equiv \sum_{\Delta\ne0} \f1\Delta \f{\v w_\Delta\c\v w_\Delta
    - \v{\m w}_\Delta\c\v{\m w}_\Delta}{N!_4},
\end{align}
\begin{align}
  W_2\p{\m h,\m w}
  \equiv \sum_{\Delta\ne0} \f1{\Delta} \sp{\f{\v w_\Delta\c\v w_\Delta
      - 2\v{\m w}_\Delta\c\v{\m w}_\Delta}{4N!_2}
    + \f{\p{3N-4}\p{\v w_\Delta\c\v w_\Delta
        - \v{\m w}_\Delta\c\v{\m w}_\Delta}}{2N!_4}},
\end{align}
where
\begin{align}
  N!_k \equiv \prod_{j=0}^{k-1} \p{N-j} = \f{N!}{\p{N-k}!}.
\end{align}
The dependence of the coefficients of $W_k\p{\m h,\m w}$ on the
interaction matrix $\m h$ is implicit in the decomposition of the
coefficient matrix $\m w$ into components $\m w_\Delta$ as
$\m w=\sum_\Delta\m w_\Delta$, where $\m w_\Delta$ is the projection
of $\m w$ onto the $\Delta$-eigenspace of the matrix $\check{\m h}$
defined in \eqref{eq:h_super_mat}.

@@

We wish to simplify
\begin{align}
  H_0 \V_2\p{\m w} \ket\psi
  = \f12 \sum_{\substack{k,\ell\\p<q}} w_{k\ell} h_{pq}
  \Pi_{pq} X_k Y_\ell \ket\psi,
  \label{eq:H_V2_psi_start}
\end{align}
which has one term with $\set{p,q}=\set{k,\ell}$, terms with
$p,q\notin\set{k,\ell}$, and terms with only one of
$p,q\in\set{k,\ell}$.  The permutation operator $\Pi_{pq}$ acts
trivially on $X_k Y_\ell \ket\psi$ when $\set{p,q}=\set{k,\ell}$ or
$p,q\notin\set{k,\ell}$, leaving a sum over $p,q$ of the form
\begin{align}
  h_{k\ell} + \sum_{\substack{p<q\\p,q\notin\set{k,\ell}}} h_{pq}
  = \sum_{p<q} h_{pq}
  - \sum_{\substack{p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  = E_0 + 2 h_{k\ell} - h_k - h_\ell
  = E_0 + 2 \bar h_{k\ell},
\end{align}
where
\begin{align}
  \bar h_{k\ell} \equiv h_{k\ell} - \f12\p{h_k + h_\ell}.
\end{align}
The corresponding terms in \eqref{eq:H_V2_psi_start} with
$p,q\notin\set{k,\ell}$ and $\set{p,q}=\set{k,\ell}$ are then
\begin{align}
  E_0 \V_2 \ket\psi
  + \sum_{k,\ell} \bar h_{k\ell} w_{k\ell} X_k Y_\ell \ket\psi.
\end{align}
Using similar techniques, we can simplify the terms in
\eqref{eq:H_V2_psi_start} with only one of $p,q\in\set{k,\ell}$ to
\begin{align}
  \f12 \sum_{k\ne\ell} \sum_q \p{h_{kq} w_{q\ell} + w_{kq} h_{q\ell}}
  X_k Y_\ell \ket\psi.
\end{align}
In total,
\begin{align}
  H_0 \V_2 \ket\psi
  &= E_0 \V_2 \ket\psi
  + \sum_{k,\ell} \bar h_{k\ell} w_{k\ell} X_k Y_\ell \ket\psi
  + \f12 \sum_{k\ne\ell} \sum_q \p{h_{kq} w_{q\ell} + w_{kq} h_{q\ell}}
  X_k Y_\ell \ket\psi \\
  &= E_0 \V_2 \ket\psi
  + \f12 \sum_{k\ne\ell} \sp{2\bar h_{k\ell} w_{k\ell}
    + \sum_q \p{h_{kq} w_{q\ell} + w_{kq} h_{q\ell}}}
  X_k Y_\ell \ket\psi
\end{align}
